Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}

### 2017-10-13 ms/ms.Rnw
release 1

% Make helvetica the default sans-serif font
% \renewcommand\sfdefault{phv}

@@

Data are currently available on GitHub and will be archived on Dryad upon publication. \\

@@

\setstretch{\stretchy}

\begin{center}
{\LARGE  \centering Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Erythranthe cardinalis}}
\end{center}

@@

% Toggle these commands to get version for doing word count
\linenumbers
% \pagenumbering{gobble}
% \raggedright

% Local adaptation and clines
Local adaptation has been documented within numerous species; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Leimu_Fischer_2008, Hereford_2009}. However, the prevalance of local adaptation remains difficult to assess because researchers rarely test for local adaptation unless there are obvious phenotypic or environmental differences (but see \citeauthor{Hereford_Winn_2008} \citeyear{Hereford_Winn_2008}). When local adaptation occurs, it frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977, Barton_1999}. Phenotypic differences between populations along a cline often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}. Despite a long history of studying local adaptation and clines, it remains challenging to identify exactly which traits are under selection and which differ for nonadaptive reasons. In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment. A related problem is identifying which of the myriad and often covarying aspects of the environment cause spatially varying selective pressures.

% Intrinsic versus plastic variation 
When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}. The traits that underlie local adaptation, however, need not mirror this pattern. Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away. Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive. We distinguish between these patterns of adaptive trait differences by referring to `genetic variation' in trait means and `genetic variation in plasticity', respectively. Genetic variation in plasticity is synonymous with genotype-by-environment interactions, or simply (G$\times$E). Genetic variation in trait means and plasticity are both involved in adaptation. For example, genetic variation in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genetic variation in plasticity cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}. In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance \citep{Flood_etal_2011}.

% Selective agents (trait-enviro assoc)
A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments. Either genetic variation in trait means and/or plasticity should vary clinally along environmental gradients. Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown. For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}. Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation. Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008, Stock_etal_2014}, and drought resistance \citep{Kooyers_etal_2015} that likely relate to climatic variation.

@@

# Legend
# pdf(paste(pathFig, "/ExperimentalDesignLegend.pdf", sep = ""), 2, 2)
par(mai = c(2, 0, 1, 0))
plot(0, 0, type = "n", xlim = c(-0.5, 1.5), ylim = c(0, 1), axes = F, xlab = "")

@@

We calculated $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$ for all two- and three-way Population $\times$ Treatment interactions for each trait. To determine whether $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$ was significantly different than 0, we calculated 95\% confidence intervals using 1000 parametric bootstrap samples simulated from fitted models. If the 95\% confidence interval for a given $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$, was greater than zero then we concluded that the Population effect size was significantly larger than the Population $\times$ Treatment effect size, and vice versa if the confidence internval was less than zero. If the confidence spanned zero, then the effect sizes are not significantly different.

@@

% This hypothesis makes several independent, testable predictions. The allocation tradeoff predicts that northern populations will provision more photosynthetic assimilate to rhizomes compared with southern populations. If southern populations are indeed more `annualized' because more frequent droughts cause mortality, then we predict that species distribution models using recent climate would best predict occurrences in the south, whereas longer term climate would be a better predictor in the north. Finally, we predict that southern populations would show greater variation in the size of recruits and higher maximum population growth rates.

>>>>

@@

Data are currently available on GitHub and Dryad \citep{Muir_Angert_2017} \\

@@

@@

% Toggle these commands to get version for doing word count
% \linenumbers
% \pagenumbering{gobble}
% \raggedright

Local adaptation has been documented within numerous species; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Leimu_Fischer_2008, Hereford_2009}. However, the prevalance of local adaptation remains difficult to assess because researchers rarely test for local adaptation unless there are obvious phenotypic or environmental differences (but see \citeauthor{Hereford_Winn_2008} \citeyear{Hereford_Winn_2008}). When local adaptation occurs, it frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977, Barton_1999}. Phenotypic differences between populations along a cline often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}. Despite a long history of studying local adaptation and clines, it remains challenging to identify exactly which traits are under selection and which differ for nonadaptive reasons. In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment. A related problem is identifying which of the myriad and often covarying aspects of the environment cause spatially varying selective pressures.

When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004}. The traits that underlie local adaptation, however, need not mirror this pattern. Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away. Alternatively, the ability to plastically respond to a particular environment or the magnitude of response to an environment could be adaptive. We distinguish between these patterns of adaptive trait differences by referring to `genetic variation in trait means' and `genetic variation in plasticity', respectively. Genetic variation in plasticity is synonymous with genotype-by-environment interactions, or simply (G$\times$E). Genetic variation in trait means and plasticity are both involved in adaptation. For example, genetic variation in photoperiod responses \citep{Blackman_etal_2011} and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genetic variation in plasticity cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}. In plants especially, we know little about the prevalence and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance \citep{Flood_etal_2011}.

A basic approach to identify candidate traits underlying local adaptation is to find associations between traits and environments. Either genetic variation in trait means and/or plasticity should vary clinally along environmental gradients. Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown. For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}. Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation. Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, leaf morphology \citep{Hopkins_etal_2008, Stock_etal_2014}, and drought resistance \citep{Kooyers_etal_2015} that likely relate to climatic variation.

@@

# Legend
par(mai = c(2, 0, 1, 0))
plot(0, 0, type = "n", xlim = c(-0.5, 1.5), ylim = c(0, 1), axes = F, xlab = "")

@@

We calculated $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$ for all two- and three-way Population $\times$ Treatment interactions for each trait. To determine whether $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$ was significantly different than 0, we calculated 95\% confidence intervals using 1000 parametric bootstrap samples simulated from fitted models. If the 95\% confidence interval for a given $\Delta \textrm{Effect Size}_{\textrm{Pop} - (\textrm{Pop} \times \textrm{Trt})}$, was greater than zero then we concluded that the Population effect size was significantly larger than the Population $\times$ Treatment effect size, and vice versa if the confidence interval was less than zero. If the confidence interval spanned zero, then the effect sizes are not significantly different.

@@
